---
title: "Assignment 2: Coding Basics"
author: Samantha Pace
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

```{r}
seq(1,30,3) # This function is creating a sequence of numbers 1 to 30 by 3
```

```{r}
sequence_by_3 <- seq(1,30,3) 
# This code chunk is indicating that the sequence generated.
#  The seq function is now assigned to the name sequence_by_3.
```

2.  Compute the mean and median of this sequence.
```{r}
mean(sequence_by_3) 
# This code chunk is referencing the object to find the mean.
```

```{r}
median(sequence_by_3) 
# This code chunk is finding the median of the object.
```

3.  Ask R to determine whether the mean is greater than the median.
```{r}
mean(sequence_by_3)>median(sequence_by_3) 
# This code chunk is asking if the mean is greater than the median. 
# Since they are both 14.5, the answer is false.
```


4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. # This function is creating a sequence of numbers 1 to 30 by 3

#2. # This code chunk is referencing the object to find the mean.

#3. # This code chunk is asking if the mean is greater than the median. 
# Since they are both 14.5, the answer is false.

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.
```{r}
students <- c('Bob', 'Joe', 'Sam', 'Corey') # This is a character vector
scores <- c(85,45,95,90) # This is a numerical vector.
outcome <- c(T, F, T, T) # This is a character vector.
passing_score <- 50

```

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.
```{r}
students_scores <- data.frame(students, scores, outcome)
print(students_scores)
```

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
print(students_scores)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different from a matrix in that it has different classes of variables. Data frames are versatile and can hold different types of data, matrices can only have a single type of tabular data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
if_else_function <- function(x){
  if(scores>passing_score){
    True
  }
    else{
      false
    }
}
try1 <-if_else_function(scores)
try1
```


```{r}
try2 <- function(x)
  ifelse(scores > passing_score, true, false)
try2(85)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Neither function worked for how I set up the code chunks, which I included above. I thought the 'ifelse' code would work better than the 'if' and 'else'.
