---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Samantha Pace"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
getwd() #checking that the EDE_Fall2023 Folder is my wd

# loading packages
library(tidyverse)
library(agricolae)
library(lubridate)

# importing data
NTL.LTER.raw <-
  read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
           stringsAsFactors = TRUE)

# fixing dates
NTL.LTER.raw$sampledate <- mdy(NTL.LTER.raw$sampledate) 

#2
# creating my theme based on the black and white theme
mytheme <-
  theme_bw(base_size = 14)+
    theme(panel.grid.major = element_line(colour = "grey80"))+
  theme(plot.title = element_text(size = rel(1)), legend.position = "top")

# setting my theme to be the default
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature recorded during July does not change with depth across all lakes.
Ha: THe mean lake temperature recorded during July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

#creating a new processed data set that is filtered to be only July, have certain columns,
# and be complete cases.
NTL.LTER.processed <-
  NTL.LTER.raw %>%
  filter(daynum > 181 & daynum < 213) %>% #filtering so it is only July
  select(lakename, year4, daynum, depth, temperature_C) %>% # selecting certain columns
  na.omit #dropping all incomplete cases

#5
#scatterplot of x=depth, y = temp limited to 35 degrees
#added titles, updated labels, and smoothed line of linear model
Depth.by.temp <-
  ggplot(NTL.LTER.processed, aes(x = depth, y = temperature_C))+
  geom_point()+
  ylim(0,35)+
  xlab("Depth")+
  ylab("Temperature (C)")+
  geom_smooth(method=lm, color = 'blue')+
  ggtitle("Temperature by Depth in July")
print(Depth.by.temp)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: As the depth increases, the temperature decreases. Depth and temperature are inversely related. Especially between depths of 2 to 6 meters, the distribution of observed temperatures ranges the widest - the distribution may be up to +/- 10 degrees from the linear model. Additionally, there are many more data points for depths under 7 meters, and much less data collected above 7 meters. The distribution of this data suggests that perhaps there is not a linear relationship between these variables.

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
# regression using lm function for temp by depth relationship 
depth.temp.regression <-
  lm(data = NTL.LTER.processed, temperature_C ~ depth)

# summary of results of regression
summary(depth.temp.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The coefficient for depth is -1.94, which tells us that when depth increases, the temperature decreases. The r-squared value is 0.7391, which means that 73.9% of the values of temperature are explained by depth. The degrees of freedom for this finding is 9720. For every one 1m increase in depth, the temperature will drop by -1.94 degrees C. The p value is less than the confidence level of 0.05, so we can determine that the coefficient is significant.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
# creating a linear model with all three potential explanatory variables, start of AIC
JulyAIC <- lm(data = NTL.LTER.processed, temperature_C ~ year4 + daynum + depth)

# running stepwise algorithim to determine set of explanatory variables
step(JulyAIC)

#10

# running multiple linear regression with year4, daynum, and depth as variables & summary
Temp.model <-
  lm(data = NTL.LTER.processed, temperature_C ~
       year4 + daynum + depth)
summary(Temp.model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests using year4, daynum, and depth to predict the temperature in the multiple regression. Removing any of these three variables would increase the AIC. The R squared value is 0.7417 so 74% of the observed variance of temperature can be explained by this model. Technically it is an increase in 0.26% higher than the model using only depth for the explanation of the observed variance. However, I would argue that this model is over-parameterized and not an improvement because adding more variables is making the model more complex, changing the degrees of freedom, and the 0.26% is not worth it.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# ANOVA models with aov function
Temp.lakename.anova <- 
  aov(data = NTL.LTER.processed, temperature_C ~ lakename)
summary(Temp.lakename.anova)

#linear models running ANOVA test
Temp.lake.anova2 <-
  lm(data = NTL.LTER.processed, temperature_C ~ lakename)
summary(Temp.lake.anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is significant difference in the mean temperature among the lakes. The estimates for each lake range between -7.4 and 17.67 (the intercept), and the p value for each lake is less than 0.05 so we can reject the null hypothesis. As a result, we reject the null hypothesis that all the mean temperatures among the lakes are the same.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, width = 2.5}
#14.
# Creating a scatterplot to look at depth and temp by lake
Plot.by.lake <-
  ggplot(NTL.LTER.processed, 
         aes(x = depth, y = temperature_C, color = lakename)) +
geom_point(alpha = 0.5) +
geom_smooth(method = "lm", se = FALSE) +
  ylim(0, 35)+
  labs(x ="Depth", y ="Temperature (C)", color = "Lakes") +
  ggtitle("Temperature by Depth in Lakes")
print(Plot.by.lake)
  
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
# Running Tukey HSD for object derived from function aov
TukeyHSD(Temp.lakename.anova)

# Groups for pairwise
Lakes.HSD.groups <- 
  HSD.test(Temp.lakename.anova, "lakename", group = TRUE)
Lakes.HSD.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Statistically speaking, Paul Lake and Ward Lake have the same mean temperature. They are all in group "c". There is not a group with only one lake in it, so there is no lake with a mean temperature that is statistically distinct from all the other lakes.

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we are looking at just Peter and Paul Lakes, we might try a two sample T-test to test the hypothesis that the means are equivalent. The alternate hypothesis would be that the the mean temperatures of Peter and Paul Lakes are statistically distinct.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# processing data to be only Crampton and Ward Lakes
Crampton.Ward.July <-
  NTL.LTER.processed %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")

# running a two sample t test:
Crampton.Ward.twosamplet <-
  t.test(Crampton.Ward.July$temperature_C ~ Crampton.Ward.July$lakename)
Crampton.Ward.twosamplet

```

>Answer: The p value of the two sample t test is 0.265, which is higher than the 0.05 confidence level, so we cannot reject the null hypothesis. Crampton and Ward Lakes, according to this test, may have the same mean temperatures. According the results found in the HSD groups in Questions 15 and 16, both Crampton and Ward Lakes were in group b, so they have mean temperatures that are not statistically distinct.
