---
title: "Assignment 4: Data Wrangling"
author: Samantha Pace
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.

The completed exercise is due on Thursday, Sept 28th @ 5:00pm.

## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.

1b. Check your working directory.

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Apply the `glimpse()` function to reveal the dimensions, column names, and structure of each dataset. 

```{r setup, message = FALSE}
#1a 
#install.packages("tidyverse") # This code chunk installs the tidyverse package.
#install.packages("lubridate") # This code chunk install lubridate package.
#install.packages("here") # This code chunk will install the here package.
# they are all now comments so it will knit.

library(tidyverse) 
library(lubridate)
library(here)
# These code chunks are loading the packages.

#1b 
getwd() # checking working directory
setwd("/home/guest/R/EDE_Fall2023") #setting wd to EDE_Fall2023
getwd() # checking wd again

#1c 
# each of these code chunks is importing one of the four datasets for EPA air 
# with string as factors set to TRUE.
EPA.O3.2018 <-
  read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE)

EPA.O3.2019 <-
  read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)

EPA.PM25.2018 <-
  read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE)

EPA.PM25.2019 <-
  read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE)

#2 
# each of these codes is applying the glimpse function to the datasets 
glimpse(EPA.O3.2018)
glimpse(EPA.O3.2019)
glimpse(EPA.PM25.2018)
glimpse(EPA.PM25.2019)
```


## Wrangle individual datasets to create processed files.

3. Change the Date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r wrangling individual datasets}
#3
# formatting each date to be a date object instead of a factor in the four data sets
EPA.O3.2018$Date <- as.Date(EPA.O3.2018$Date, format = "%m/%d/%Y")
EPA.O3.2019$Date <- as.Date(EPA.O3.2019$Date, format = "%m/%d/%Y")
EPA.PM25.2018$Date <- as.Date(EPA.PM25.2018$Date, format = "%m/%d/%Y")
EPA.PM25.2019$Date <- as.Date(EPA.PM25.2019$Date, format = "%m/%d/%Y")

#4
# selecting specific columns in the imported data to be included in new data subsets.
EPA.O3.2018.selected <-
  select(EPA.O3.2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, 
         SITE_LATITUDE, SITE_LONGITUDE )

EPA.O3.2019.selected <-
  select(EPA.O3.2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, 
         SITE_LATITUDE, SITE_LONGITUDE )

EPA.PM25.2018.selected <-
  select(EPA.PM25.2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, 
         SITE_LATITUDE, SITE_LONGITUDE )

EPA.PM25.2019.selected <-
  select(EPA.PM25.2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, 
         SITE_LATITUDE, SITE_LONGITUDE )
#5
# using the mutate function to insert "PM2.5" into the AQS_PARAMETER_DESC column for
# both PM2.5 datasets

EPA.PM25.2018.selected <- 
  mutate(EPA.PM25.2018.selected, AQS_PARAMETER_DESC = "PM2.5")
EPA.PM25.2019.selected <- 
  mutate(EPA.PM25.2019.selected, AQS_PARAMETER_DESC = "PM2.5")

#6
# each of the four code chunks below are writing a new csv file for each of the 
# processed air quality datasets into the processed data folder.
write.csv(EPA.O3.2018.selected, row.names = FALSE, 
          file = "./Data/Processed/EPAair_O3_NC2018_Processed.csv")

write.csv(EPA.O3.2019.selected, row.names = FALSE, 
          file = "./Data/Processed/EPAair_O3_NC2019_Processed.csv")

write.csv(EPA.PM25.2018.selected, row.names = FALSE, 
          file = "./Data/Processed/EPAair_PM25_NC2018_Processed.csv")

write.csv(EPA.PM25.2019.selected, row.names = FALSE, 
          file = "./Data/Processed/EPAair_PM25_NC2019_Processed.csv")

```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include only sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r combining, pipe, processing, saving}
#7 

# this rbind function is combining all four of the processed individual datasets
rbind.four.selected.datasets <- 
  rbind(EPA.O3.2018.selected, EPA.O3.2019.selected, 
        EPA.PM25.2018.selected, EPA.PM25.2019.selected)

#8 
# This pipe is creating a new dataset All.O3.PM25 that filters, groups, summarises, and 
# mutates.
All.O3.PM25 <-
  rbind.four.selected.datasets %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", 
                          "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", 
                          "West Johnston Co.", "Garinger High School", "Castle Hayne", 
                          "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE), 
            meanlat = mean(SITE_LATITUDE),
            meanlong = mean(SITE_LONGITUDE)) %>%
  mutate(Month = month(Date), Year = year(Date))

#9
# this code is doing a spread such that ozone and PM2.5 are their own columns now, replacing
# parameter and mean AQI

All.O3.PM25.spread1 <-
  pivot_wider(All.O3.PM25, names_from = AQS_PARAMETER_DESC, values_from = meanAQI)
  

#10
# Checking the dimensions of the spread
dim(All.O3.PM25.spread1)
  
#11
# writing a new CSV and saving it into the processed data folder.
write.csv(All.O3.PM25.spread1, row.names = FALSE, 
          file ="Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r summary}

#12
 # This code chuck is making a summary of the pivot wider data subset, grouping by
# date, site.name, month, and year. There is a summary line and all NA's are dropped from the ozone column.

EPAair.summary <-
  All.O3.PM25.spread1 %>%
  group_by(Date, Site.Name, Month, Year) %>%
  summarise(mean.Ozone = mean(Ozone), 
            mean.PM25 = mean(PM2.5)) %>%
  drop_na(mean.Ozone)

#13
# determining the dimensions of the summary dataset
dim(EPAair.summary)

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: Since we want to keep the rows that have NA in the PM25 column, we don't want to omit all the NAs in the dataset, which is what the na.omit function does. The na.omit will drop entire rows with any NA in them. The drop_na function will only look at the column we specify, in this case it is the ozone column. This allows us to keep the NAs in the PM25 column.
